Demographic Shifts in Toronto's Homeless Population: Evaluating and Reassessing City Solutions

### 2024-09-27 paper/paper.qmd
final edition

While the city has made progress in expanding housing, we find that a housing-centric approach may not be sufficient to address the needs of the evolving homeless population. Our analysis reveals significant demographic changes in Toronto’s homeless population over the past six years, including increases in the proportion of chronically homeless individuals, refugees, and working-age adults. These demographic shifts suggest that the traditional emphasis on housing development alone is no longer be an adequate solution.

In this paper, we explore these demographic changes using monthly data from Toronto's Shelter Management Information System (SMIS) from January 2018 to June 2024 [@toronto_shelter_data]. We detail the data collection process, cleaning methods, and key findings in the Data section. In the final section, Discussion, we assess the implications of our results for policy and propose recommendations for more comprehensive strategies that extend beyond housing solutions to address the diverse needs of Toronto's homeless population.

@@

While other datasets, such as daily shelter occupancy, are available, they lack the demographic depth necessary for this study. Data loading, cleaning, and analysis were carried out in **R** [@citeR], with packages **tidyverse** [@tidy], **here** [@here], **styler** [@style], and **knitr** [@knitr]. This comprehensive approach ensures that our analysis accurately reflects the realities faced by individuals experiencing homelessness in Toronto.

@@

### Chronic trend
@fig-Chronic shows the trend of the proportion of chronic population among all homeless people over the quarters of each year. Here, according to [@chronic], chronic homelessness refers to "People who meet one of the two following criteria, as per the federal definition of chronic homelessness. The person has recorded a minimum of 180 overnight stays in the past year (365 days), or the person has recurrent overnight stays over the past three years with a cumulative duration of at least 546 nights." We can see from the graph that the proportion of chronic homeless increased a lot, from less than 35% in the first quarter of 2018 to more than 65% in the second quarter of 2024. 
```{r}
#| label: fig-Chronic
#| echo: false
#| fig-cap: ""

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    x = "Quarters of a year", y = "Proportion",
    title = "Proportion of chronic homeless population to toal homeless\npopulation in Toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Age trend
@fig-WorkingAge displays the trend of the proportion of different working age statuses among the actively homeless population over the quarters of each year. Here, the working age statuses are categorized into three groups: Early Career Stage, Working Age, and Pre-Retirement and Retirement Age. Even though the trend is not dramatic, We can see from the graph that the proportion of the Early Career Stage population as well as pre-retire and retired has decreased (from 29% and 17% separately in the first quarter of 2018 to 24.5% and 14.5% in the second quarter in 2024). On the other hand, the working age proportion increased to 61% in 2024Q2 from 54% in 2018Q1. This indicates the growing population who are in the working age cannot afford their housing. 
```{r}
#| label: fig-WorkingAge
#| echo: false
#| fig-cap: ""

@@

First, as the result section points out, the proportion of the chronically homeless population nearly doubled in the past six years. Despite its nature of long-lasting homeless status, studies have also shown that chronic homelessness is closely associated with chronic illness or other mental issues @chronic. Research further indicates that the average income of people experiencing mental illness and different types of disorders is significantly lower than normal people's income @Sareen2011. This makes it particularly difficult for people who have already experienced health problems to escape homelessness. In addition, homelessness exacerbates health problems due to lack of adequate food, discrimination, barriers to health care, etc. [@Sleet2021]. A combination of these would become a harmful cycle: illness leads to a decrease in income, which leads to becoming homeless, which would further worsen individual illness.

>>>>

While the city has made progress in expanding housing, we find that a housing-centric approach may not be sufficient to address the needs of the evolving homeless population. Our analysis find significant demographic changes in Toronto’s homeless population over the past six years, including increases in the proportion of chronically homeless individuals, refugees, and working-age adults. These demographic shifts indicate that the traditional emphasis on housing development alone is no longer be an adequate solution.

In this paper, we explore these demographic changes using monthly data from Toronto's Shelter Management Information System (SMIS) from January 2018 to June 2024 [@toronto_shelter_data]. We detail the data collection process, cleaning methods, and key findings in the **Data** section. In the final section, **Discussion**, we assess the implications of our results for policy and propose recommendations for more comprehensive strategies that extend beyond housing solutions to address the diverse needs of Toronto's homeless population.

@@

While other datasets, such as daily shelter occupancy, are available, they lack the demographic information necessary for this study. Data loading, cleaning, and analysis were carried out in **R** [@citeR], with packages **tidyverse** [@tidy], **here** [@here], **styler** [@style], and **knitr** [@knitr].

@@

### Chronic trend
@fig-Chronic shows the trend of the proportion of chronic population among all homeless people over the quarters of each year. Here, according to [@chronic], chronic homelessness refers to "People who meet one of the two following criteria, as per the federal definition of chronic homelessness. The person has recorded a minimum of 180 overnight stays in the past year (365 days), or the person has recurrent overnight stays over the past three years with a cumulative duration of at least 546 nights." We can see from the graph that the proportion of chronic homeless nealy doubled, from less than 35% in the first quarter of 2018 to more than 65% in the second quarter of 2024. 
```{r}
#| label: fig-Chronic
#| echo: false
#| fig-cap: ""

@@

# Create a bar graph
ggplot(plot_data, aes(x = Quarter, y = Proportion)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    x = "Quarters of a year", y = "Proportion",
    title = "Proportion of chronic homeless population to toal homeless\npopulation in Toronto, measured quarterly"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Age trend
@fig-WorkingAge displays the trend of the proportion of different working age statuses among the actively homeless population over the quarters of each year. Even though the trend is not dramatic, We can see from the graph that the proportion of the Early Career Stage population as well as pre-retire and retired has decreased (from 29% and 17% separately in the first quarter of 2018 to 24.5% and 14.5% in the second quarter in 2024). On the other hand, the working age proportion increased to 61% in 2024Q2 from 54% in 2018Q1. This indicates the growing population who are in the working age cannot afford their housing. 
```{r}
#| label: fig-WorkingAge
#| echo: false
#| fig-cap: ""

@@

First, as the result section points out, the proportion of the chronically homeless population increased 30% in the past six years. Despite its nature of long-lasting homeless status, studies have also shown that chronic homelessness is closely associated with chronic illness or other mental issues [@chronic]. Research further indicates that the average income of people experiencing mental illness and different types of disorders is significantly lower than normal people's income [@Sareen2011]. This makes it particularly difficult for people who have already experienced health problems to escape homelessness. In addition, homelessness exacerbates health problems due to lack of adequate food, discrimination, barriers to health care, etc. [@Sleet2021]. A combination of these would become a harmful cycle: illness leads to a decrease in income, which leads to becoming homeless, which would further worsen individual illness.
